---
title: "Homework 03: Use dplyr/ggplot2 to manipulate and explore data"
author: "Aidan Hughes"
date: "October 1, 2018"
output:
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
***
# Overview

We were given a list of tasks to choose three from and complete. Here are my choices:

* *"How is life expectancy changing over time on different continents?"*

* *"Get the maximum and minimum of GDP per capita for all continents."*

* *"Report the absolute and/or relative abundance of countries with low life expectancy over time by continent."*


Don't forget to include our packages:

```{r packages}
suppressPackageStartupMessages(library("gapminder"))
suppressPackageStartupMessages(library("tidyverse"))
```



# Task 1: *"How is life expectancy changing over time on different continents?"*

To see how life expectancy over time, let's find the mean life expectancy of each continent for each year we have data for. We'll print a sample of the data in a table since printing all the data would be a little overwhelming.

```{r lifeExp each year}
lifeExpOverTime <- gapminder %>%
  select(year, continent, lifeExp) %>%
  group_by(continent, year) %>%
  mutate(meanLifeExp = mean(lifeExp), sdLifeExp = sd(lifeExp)) %>%
  select(year, continent, meanLifeExp, sdLifeExp)

# Sample a single row for each of the continents
lifeExpOverTime %>%
  ungroup() %>%
  distinct(continent, .keep_all = TRUE) %>%
  knitr::kable()
```

A simple line plot of mean life expectancy vs. year will give us an idea of what's happening over time.

```{r line plot}
lifeExpPlot <- ggplot(lifeExpOverTime, aes(y = meanLifeExp, x = year, color = continent, group = continent)) +
  ggtitle("Life Expectancy Vs. Year") +
  
  #Center the title, left-aligned by default
  theme(plot.title = element_text(hjust = 0.5))

lifeExpPlot +
  geom_line()
```

Plotting the mean only tells us part of the story. What about the standard devation of the life expectancy? We can try representing this as error bars.

```{r line with error bars}
lifeExpPlot +
  geom_line() +
  geom_errorbar(aes(ymin = meanLifeExp - sdLifeExp,
                    ymax = meanLifeExp + sdLifeExp))
```

The plot is getting a little messy now. A ribbon plot can tell us the same information with less clutter.

```{r ribbon plot}
lifeExpPlot +
  geom_ribbon(aes(ymin = meanLifeExp - sdLifeExp,
                  ymax = meanLifeExp + sdLifeExp,
                  fill=continent), alpha=0.2)
```



# Task 2: *"Get the maximum and minimum of GDP per capita for all continents."*

Start by searching through all the data to find the minimum and maximum GDP per capita of each continent.

```{r GDP per cap min max}
gdpPerCapMaxMin <- gapminder %>%
  select(continent, gdpPercap) %>%
  group_by(continent) %>%
  mutate(mingdpPerCap = min(gdpPercap), maxgdpPerCap = max(gdpPercap)) %>%
  distinct(continent, mingdpPerCap, maxgdpPerCap)

knitr::kable(gdpPerCapMaxMin)
```

We can go further and find which country in each continent had the minimum and maximum GDP per capita, along with which year.

```{r MaxMin GDP countries}
countries <- gapminder %>%
  select(country, year, continent, gdpPercap) %>%
  filter(gdpPercap %in% gdpPerCapMaxMin$mingdpPerCap |
         gdpPercap %in% gdpPerCapMaxMin$maxgdpPerCap) %>%
  mutate(MinOrMax = ifelse(gdpPercap %in% gdpPerCapMaxMin$mingdpPerCap, "min", "max")) %>%
  arrange(continent, desc(MinOrMax))

knitr::kable(countries)
```

We can plot the max and min GDP per capita of each continent conveniently in a bar graph.

```{r bar plot}
ggplot(countries, aes(x = continent, y = gdpPercap, fill = MinOrMax)) +
  geom_col(position = "dodge") +
  
  # Labelling
  geom_text(aes(label = country, vjust = -0.5), size = 2, fontface="bold") +
  labs(y = "GDP Per Capita", x = "Continent", fill = "min. or max. \n GDP") +
  
  # ggplot automatically adds a legend for size and alpha, need to remove them
  guides(size=FALSE, alpha=FALSE) +
  
  ggtitle("Maximum and Minimum \nGDP Per Capita") +
  
  # Center the title, left-aligned by default
  theme(plot.title = element_text(hjust = 0.5))
```

# Task 3: *"Compute a trimmed mean of life expectancy for different years."*

Let's calculate the trimmed mean using the `summarize` function, and see how the data changes depending on the trim value.

```{r trimmed means}
trims <- gapminder %>%
  group_by(year) %>%
  summarize(trim0.1 = mean(lifeExp, trim=0.1),
            trim0.3 = mean(lifeExp, trim=0.3),
            trim0.5 = mean(lifeExp, trim=0.5)) %>%
  gather(key = "Trim", value = "Mean", contains("trim"))
```
We'll plot a sample of the data from each of the trimmed means.

```{r}
trims %>%
  filter(year < 1968) %>%
knitr::kable()
```


Let's use a counts plot to visualize the data.

```{r counts plot}
trims %>%
  ggplot(aes(x = Trim, y = Mean)) +
  geom_count(col="blue", show.legend=FALSE, alpha=0.5) 
```

Not quite enough data points to make a great counts plot, but still shows what's happening when we trim the mean.

# That's all, thanks for your time! :smile:
