---
title: "hw03-aidanh14"
author: "Aidan Hughes"
date: "October 1, 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Homework 03: Use dplyr/ggplot2 to manipulate and explore data

We were given a list of tasks to choose three from and complete. Here are my choices:

* *"How is life expectancy changing over time on different continents?"*

* *"Task 2"*

* *"Task 3"*


Don't forget to include our packages:

```{r packages}
suppressPackageStartupMessages(library("gapminder"))
suppressPackageStartupMessages(library("tidyverse"))
```



## "How is life expectancy changing over time on different continents?"

To see how life expectancy over time, let's find the mean life expectancy of each continent for each year we have data for.

```{r lifeExp each year}
lifeExpOverTime <- gapminder %>%
  select(year, continent, lifeExp) %>%
  group_by(continent, year) %>%
  mutate(meanLifeExp = mean(lifeExp), sdLifeExp = sd(lifeExp)) %>%
  select(year, continent, meanLifeExp, sdLifeExp)

lifeExpOverTime
```

A simple line plot of mean life expectancy vs. year will give us an idea of what's happening over time.

```{r line plot}
lifeExpPlot <- ggplot(lifeExpOverTime, aes(y = meanLifeExp, x = year, color=continent, group=continent)) +
  ggtitle("Mean Life Expectancy Vs. Year") +
  theme(plot.title = element_text(hjust = 0.5))

lifeExpPlot +
  geom_line()
```

Plotting the mean only tells us part of the story. What about the standard devation of the life expectancy? We can try representing this as error bars.

```{r line with error bars}
lifeExpPlot +
  geom_line() +
  geom_errorbar(aes(ymin = meanLifeExp - sdLifeExp, ymax = meanLifeExp + sdLifeExp))
```

The plot is getting a little messy now. A ribbon plot can tell us the same information with less clutter.

```{r ribbon plot}
lifeExpPlot +
  geom_ribbon(aes(ymin = meanLifeExp - sdLifeExp, ymax = meanLifeExp + sdLifeExp, fill=continent), alpha=0.2)
```

Now *that's* a clean plot that gives us all the information we need.

