---
title: "Homework 03: Use dplyr/ggplot2 to manipulate and explore data"
author: "Aidan Hughes"
date: "October 1, 2018"
output:
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We were given a list of tasks to choose three from and complete. Here are my choices:

* *"How is life expectancy changing over time on different continents?"*

* *"Get the maximum and minimum of GDP per capita for all continents."*

* *"Task 3"*


Don't forget to include our packages:

```{r packages}
suppressPackageStartupMessages(library("gapminder"))
suppressPackageStartupMessages(library("tidyverse"))
```



## Task 1: "How is life expectancy changing over time on different continents?"

To see how life expectancy over time, let's find the mean life expectancy of each continent for each year we have data for. We'll print a sample of the data in a table since printing all the data would be a little overwhelming.

```{r lifeExp each year}
lifeExpOverTime <- gapminder %>%
  select(year, continent, lifeExp) %>%
  group_by(continent, year) %>%
  mutate(meanLifeExp = mean(lifeExp), sdLifeExp = sd(lifeExp)) %>%
  select(year, continent, meanLifeExp, sdLifeExp)

# Sample a single row for each of the continents
lifeExpOverTime %>%
  ungroup() %>%
  distinct(continent, .keep_all = TRUE) %>%
  knitr::kable()
```

A simple line plot of mean life expectancy vs. year will give us an idea of what's happening over time.

```{r line plot}
lifeExpPlot <- ggplot(lifeExpOverTime, aes(y = meanLifeExp, x = year, color = continent, group = continent)) +
  ggtitle("Life Expectancy Vs. Year") +
  
  #Center the title, left-aligned by default
  theme(plot.title = element_text(hjust = 0.5))

lifeExpPlot +
  geom_line()
```

Plotting the mean only tells us part of the story. What about the standard devation of the life expectancy? We can try representing this as error bars.

```{r line with error bars}
lifeExpPlot +
  geom_line() +
  geom_errorbar(aes(ymin = meanLifeExp - sdLifeExp,
                    ymax = meanLifeExp + sdLifeExp))
```

The plot is getting a little messy now. A ribbon plot can tell us the same information with less clutter.

```{r ribbon plot}
lifeExpPlot +
  geom_ribbon(aes(ymin = meanLifeExp - sdLifeExp,
                  ymax = meanLifeExp + sdLifeExp,
                  fill=continent), alpha=0.2)
```



## Task 2: "Get the maximum and minimum of GDP per capita for all continents."

Start by searching through all the data to find the minimum and maximum GDP per capita of each continent.

```{r GDP per cap min max}
gdpPerCapMaxMin <- gapminder %>%
  select(continent, gdpPercap) %>%
  group_by(continent) %>%
  mutate(mingdpPerCap = min(gdpPercap), maxgdpPerCap = max(gdpPercap)) %>%
  distinct(continent, mingdpPerCap, maxgdpPerCap)

knitr::kable(gdpPerCapMaxMin)
```

We can go further and find which country in each continent had the minimum and maximum GDP per capita, along with which year.

```{r MaxMin GDP countries}
countries <- gapminder %>%
  select(country, year, continent, gdpPercap) %>%
  filter(gdpPercap %in% gdpPerCapMaxMin$mingdpPerCap |
         gdpPercap %in% gdpPerCapMaxMin$maxgdpPerCap) %>%
  mutate(MinOrMax = ifelse(gdpPercap %in% gdpPerCapMaxMin$mingdpPerCap, "min", "max")) %>%
  arrange(continent, desc(MinOrMax))

knitr::kable(countries)
```

We can plot the max and min GDP per capita of each continent conveniently in a bar graph.

```{r bar plot}
ggplot(countries, aes(x = continent, y = gdpPercap, fill = MinOrMax)) +
  geom_col(position = "dodge") +
  
  # Labelling
  geom_text(aes(label = country, vjust = -0.5), size = 2, fontface="bold") +
  labs(y = "GDP Per Capita", x = "Continent", fill = "min. or max. \n GDP") +
  
  # ggplot automatically adds a legend for size and alpha, need to remove them
  guides(size=FALSE, alpha=FALSE) +
  
  ggtitle("Maximum and Minimum \nGDP Per Capita") +
  
  # Center the title, left-aligned by default
  theme(plot.title = element_text(hjust = 0.5))
```

## Task 3: 